---
title: "ENVS 193DS Final"
format: pdf
editor: source
---

[Github repository](https://github.com/lucia-murillo/ENVS-193DS_spring-2026_final)

```{r}
#| label: set-up

# reading in packages 
library(tidyverse)
library(janitor)
library(here)
library(ggeffects)
library(MuMIn)
library(dplyr)
library(rstatix)

# creating new object from csv in `data` folder  
# using "here" to identify location of file   
nests <- read_csv(here("data", "nest_data_final.csv"))

# creating new value for removed variables in locations column
loc_removed <- c("Cafe", "Amtrak", 
        "Boyfriend's house", "UCen", 
        "Arbor", "SRB")
# creating new object from personal data 
# # using "here" to identify location of file   
personal <- read_csv(here("data",
                          "personal.csv")) |> 
  # cleaning columnn names 
  clean_names() |> 
  # removing loctations with fewer than 5 observations 
 filter(!(location %in% loc_removed))

```
** make invisible!! ** 

# Problem 1: Research writing 

## (a) 
In part 1 they used a linear model.  
In part 2 they used a Pearson's test. 

## (b) 
We found that the distance to the water's edge tends to predict the probability that American Avocets would use a habitat. *** 

We also found a relationship between bird species (American Avocet, Forster’s Tern, Black-necked Stilt) and the type of habitat (managed pond, non-tidal marsh, salt pond, tidal marsh, or tidal mudflat). ****

## (c) 
One additional variable that could influence the probability of American Avocet microhabitat use is the type of habitat, which is a categorical variable. They may have preferred food sources or shelter in some habitats and not others, which could influence the probability of microhabitat use. This was a relationship that was already explored ****** maybe not since alr explored?? 

# Problem 2: Data analysis 

```{r}
#| label: data-exp

```
**MAKE SURE TO MAKE INVISIBLE** 

## (a)
```{r}
#| label: clean-and-wrangle
# creating new object from nests dataframe
nests_clean <- nests |> 
  # creating new column with full names of ants 
  mutate(species_name = recode_values(
    ant_species,
  "RRB" ~ "C. mimosae",
  "AB" ~ "C. sjostedti",
  "BBR" ~ "C. nigriceps"
    )) |> 
  # reordering species levels 
  mutate(species_name = 
           as_factor(species_name),
         species_name = 
           fct_relevel(species_name, 
                       "C. sjostedti",
                       "C. mimosae",
                       "C. nigriceps"
                       )) |> 
  # selecting columns of interest (species name, case_control, and height_cm)
  select(species_name, case_control, height_cm) |> 
  # dropping N/A values from species name column 
  drop_na(species_name)
```

```{r}
#| label: checking-structure
# checking structure of data frame 
str(nests_clean)
```
** MAKE SURE STRUCTURE LOOKS GOOD! ** 

```{r}
#| label: ten-rows
# displaying from nests_clean 
slice_sample(
  nests_clean,
  # displaying 10 rows 
  n = 10)
```

## (b) 
In this data set, the 0s and 1s represent whether a tree contained a nest (1) or was an "available" tree for comparison (0). 
**expand on this?** 

## (c) 

Tree height is a continuous variable. I would expect that as height increases, the probability of nesting would also increase. This is supported by the article, which says that for every 1m increase in height, the odds that birds nested in a tree increased by 20% (Results section, paragraph 2). 

Acacia ant species are categorical variables. I would expect that in trees with C. nigriceps and C. mimosae, there would be a higher probability of nest occurrence. Based on the article, it seems that these ants are the most aggressive. This seems to increase probability of nests because these ants may help reduce the risk of nest predation (Discussion section of the article in paragraph 1). 

## (d) 
| Model number | Tree height | Ant species | Interaction term |
|--------------|:-----------:|:-----------:|:---------------------------------:|  
| 0            |  X         |  X           | No               |
| 1            |  X          |  X          | Yes              |
## (e) 
```{r}
#| label: model-fitting

# model 0: additive model 
model1 <- lm(
  # formula: species and height
  case_control ~ species_name + height_cm,
  # from clean data frame 
  data = nests_clean
)
# model 2: interactive model 
model2 <- lm(
  # formula: height interacting with species
  case_control ~ height_cm * species_name,
  # from clean data frame 
  data = nests_clean)

```
** make sure to hide **!! 

## (f) 
```{r}
# calculating AIC 
AICc(
  model1,
  model2 
) |> 
  # arranging by AIC 
  arrange(AICc)
```

The best model as determined by Akaike’s Information Criterion (AIC) is the one that uses both ant species and tree height to predict nest occurance (the full model). 

## (g) 

```{r}
#| label: checking-diagnostics

# displaying plots in 2x2 grid 
par(mfrow = c(2, 2))
# 
plot(model1)
```


## (h) 

```{r}
#| label: visual



```
# from woekshop 
ggplot(data = nests_clean,
       mapping = aes(x = species_name, 
                     y = case_control,
                     color = species_name)) + 
  geom_jitter(height = 0, 
              width = 0.1,
              alpha = 0.8) +
  geom_pointrange(data = preds,
                  mapping = aes(x = species_name,
                               y = predicted,
                               ymin = conf.low,
                               ymax = conf.high)) + 
  facet_wrap(~species_name) + 
  theme(legend.position = "none")

## (i)

## (j)

```{r}
#| label: model-preds

```
# workshop data 
preds <- ggpredict(
  model4,
  terms = c("water_treatment", "species_name")
) |> 
  rename(water_treatment = x,  
         species_name = group)

## (k)

# Problem 3: Affective and exploratory visualizations

## (a) 

The exploratory visualizations I made in Homework 2 are very different from the affective visualization I made for Homework 3. The exploratory visualizations were made to get information across in a simple and understandable way. The affective visualization contains data, but that is not the only goal. It is more about  being visually interesting with the explanation of the data being represented being on another sheet of paper. 

They are all similar in that they come from the same dataset. The affective visualization is also similar to the boxplot from the exploratory visualization. They bot display the productivty at different locations. 

In the affective visualization, the mean productivity levels are displayed, which show a similar pattern to the boxplot, which shows the median productivity among other things. The productivity based on the time of start does not share any patterns with the other two that I can determine. 

In week 9, my group thought my idea was good. However, some members suggested that I add other variables from my data. I did not end up doing this because I was unsure how to do that in a way that would make sense logically and artistically. 

## (b) 

I will do a Kruskall-Wallis test to compare the median productivity rank at each location. It will tell me if there is a significant difference in my productivity levels in the library compared to at home. Since my predictor variable (location) is two unordered groups, it will work well for this test. My response variable (productivity) is ordinal, which works well for a Kruskall-Wallis test since it compares the ranks of scores and does not assume a normal distribution. Each study session is an independent observation. 

## (c) 

### Checking assumptions 
1. Independence 
Each study session is an independent observation. 
2. Sample size: each group should have at least 5 observations. 
Locations that had fewer than 5 observations were removed, leaving "Home" and "Library" 
```{r}
personal |> 
  count(location, sort = TRUE)
```

```{r}
dunn_test(
  productivity_rating_1_5 ~ location,
  data = personal,
  p.adjust.method = "bonferroni"
)
```


### Running analysis 

```{r}
#| label: k-wallis

# running kruskal-wallis test 
kruskal.test(
  # productivty as a function of location 
  productivity_rating_1_5 ~ location, 
  # data: personal data 
  data = personal 
)
```

```{r}
#| label: medians
# starting with personal data
personal |> 
  # grouping by location 
  group_by(location) |> 
  # calculating median productivity for each location 
  summarise(median_prod = 
              median(productivity_rating_1_5))
```


### Effect size 

```{r}
#| label: effect-size

# determining effect size  
kruskal_effsize(
  # productivity rating as a function of location 
  productivity_rating_1_5 ~ location, 
  # data: personal data 
  data = personal)
```

## (d)
```{r}
#| label: boxplots
# base layer ggplot
# data: personal data 
ggplot(data = personal,
       # x-axis: location
       # y-axis: productivity rating 
       mapping = aes(x = location,
                     y = productivity_rating_1_5,
                     # color by location 
                     color = location)) + 
  # first layer: boxplot 
  geom_boxplot() + 
  # second layer: jitter plot 
  # adjusted height, width, opacity of points   
  geom_jitter(height = 0,
              width = 0.1,
              alpha = 0.8) + 
  # changing colors 
  scale_color_manual(values = c("purple3", "navy")) + 
  # changing axis titles and adding a title 
  labs(x = "Location", 
       y = "Productivity rating (1-5)", 
       title = "Median productivity is higher in the library") + 
  # minimal theme 
  theme_minimal() 

```

## (e)
**Figure 1: Comparison of median productivity levels at home and the library.** Each boxplot represents a location where I have studied, and each point represents a single study session. The bold line in the boxes represent the median value, and the top and bottom of the boxes represent the 75th and 25th percentiles respectively. The whiskers span the distance of the lowest to the highest 25% of the data.

## (f)
I found a small (η2 = 0.03)
difference in median productivity 
between the library and my home (Kruskal-Wallis test,
$\chi^2$(1)= 0.234, p = 0.629, ⍺ = 0.05).
There was not a significant difference in median productivity between the library (median productivity = 5) and my home (median productivity = 4.5) (Dunn's post hoc test: adjusted p-value= 0.629). 
